---
title: "Distributed IEM"
author: "Bronwyn McCall"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# IEM algorithm:
1. Have initial values for local and global sufficient statistics. (and parameters).
2. Have data stored on different nodes.
3. Start at node one and calculate the local sufficient statistics.
    3.1 update the local sufficient statistics 
    3.2 upate the global sufficient statistics
4. Update the parameter estimates
5. Check convergence. If not met, then send to the next node.

```{r libaries}
library(ggplot2)
```
## Generate the data
```{r data}
set.seed(123)

# SET MODEL HYPERPARAMETERS AND SETTINGS
K <- 2 # number of components
n <- 1000 # sample size
m <- 4 # number of nodes
n_m <- n/m # sample size per node

# TRUE PARAMETERS
pi_1 <- 0.3
pi_2 <- 1 - pi_1
mu_1 <- 0
mu_2 <- 2
si_1 <- 1
si_2 <- 2
probs <- c(pi_1, pi_2)
means <- c(mu_1, mu_2)
stds <- c(si_1, si_2)
params <- list("mixing_probs" = probs, "means" = means, "stds" = stds)

# PARAMETER STARTING VALUES
pi_1_init <- 0.5
pi_2_init <- 0.5
mu_1_init <- 0.2
mu_2_init <- 1.5
si_1_init <- 1
si_2_init <- 1.5
probs_init <- c(pi_1_init, pi_2_init)
means_init <- c(mu_1_init, mu_2_init)
stds_init <- c(si_1_init, si_2_init)
params_init <- list("mixing_probs" = probs_init,
                    "means" = means_init,
                    "stds" = stds_init)

# GENERATING DATA
generate_data <- function(means, stds, probs, n){
  ## Ensure different parameter lengths match
    if(length(means) != length(stds) || length(means) != length(probs)){
        stop("Number of means, standard deviations and components probabilities 
             must be equal.")
  }
    data <- matrix(nrow = n)
    for(i in 1:n){
        component <- rmultinom(1, 1, probs) + 1
        data[i] <- rnorm(1, mean = means[component], sd = stds[component])
  }
    return(data)
}

data_nodes <- lapply(1:m, function(x) generate_data(means, stds, probs, n_m))

# PLOT DATA
plot_data_GMM <- function(data, n_components, params, overlay = TRUE,
                      freq = FALSE){
  ## Extracting parameters
  means <- params$means
  stds <- params$stds
  probs <- params$mixing_probs
  ## Histogram
  hist(data, breaks = 50, col = "lightblue", xlab = "Value", 
       main = "Histogram of two-component univariate Gaussian mixture model",
       freq = freq)
  ## Overlayed line plot
  if (overlay) {
    colors <- rainbow(n_components)
    for (i in 1:n_components){
      curve(dnorm(x, mean = means[i], sd = stds[i])*probs[i], add = TRUE, 
            col = colors[i], lwd = 2)
    }
  }
}


## Plots
plot_data_GMM(data = data_nodes[[1]], K, params) # data node 1
plot_data_GMM(data = data_nodes[[2]], K, params) # data node 2
plot_data_GMM(data = data_nodes[[3]], K, params) # data node 3
plot_data_GMM(data = data_nodes[[4]], K, params) # data node 4
plot_data_GMM(data = unlist(data_nodes), K, params) # all data
```

## Functions to get local sufficient statistics

Functions:

1. calculate_data_belonings
2. calculate_node_local_sufficient_statistics

### Calculate data belongings
* Used on each worker node
* Calculates the data points conditional component belongings (probabilities) based on the data and the current parameter estimates
* For each data point $\sum_k^K \pi_k = 1$ and $\pi_k \in [0, 1]$
```{r data belongings}
calculate_data_belongings <- function(params, data){

  
  ### Extracting parameters and checking if they are they same.
  means <- params$means
  stds <- params$stds
  probs <- params$mixing_probs
  if (length(means) != length(stds) || length(means) != length(probs)){
    stop("Number of means, standard deviations, and weights must be equal.")
  }
  
  ### Get the data sample and number of components
  n_components <- length(means) 
  n <- length(data)
  
  ### Calculating the belongings 
  gamma_num <- matrix(nrow = n, ncol = n_components)
  for(i in 1:n_components){
    gamma_num[,i] <- probs[i] * dnorm(data, mean = means[i], sd = stds[i])
  }
  gamma_den <- rowSums(gamma_num)
  gamma <- gamma_num / gamma_den
  
  ### Check if data belongings sum to 1 (have grace for rounding errors).
  if (any(abs(rowSums(gamma) - 1) > .Machine$double.eps * n)){
      stop("Calculated belongings for each data point do not sum to 1.")
  }
  
  ### Check if any data belonging is negative
  if (any(gamma < 0)){
      stop("Calculated belongings contain negative values.")
  }
  return(gamma)
}

### test (vectorisation)
# lapply(1:m, function(x) calculate_data_belongings(params = params_init, data = data_nodes[[x]]))

### test (IEM)
for(i in 1:m){
    belongings = calculate_data_belongings(params = params_init, data = data_nodes[[i]])
    print(belongings)
}
```

### calculate_node_local_sufficient_statistics

* For each node it calculates the nodes respective local sufficient statistics
\begin{equation}
\begin{split}
a_{m,k}^t & = \sum_{i=1}^{N_m} \gamma_{m,i,k}^{t} \\
b_{m,k}^t & = \sum_{i=1}^{N_m} \gamma_{m,i,k}^{t}y_{m,i} \\
c_{m,k}^t & = \sum_{i=1}^{N_m} \gamma_{m,i,k}^{t}y_{m,i}^2 
\end{split}
\end{equation}

```{r node local statistics}
calculate_node_local_sufficient_statistics <- function(data, params){
    ### Calculating the data belongings
    belongings <- calculate_data_belongings(params = params, data = data)
    
    ### Calculating the local sufficient statistics for the node
    #### Number 1: sum of data belongings
    sufficient_stat_one <- colSums(belongings)
    
    #### Number 2: sum of product (data and data belongings)
    sufficient_stat_two <- colSums(belongings * matrix(data, nrow = length(data), ncol = ncol(belongings), byrow = FALSE))
    
    #### Number 3: sum of product (data squared and data belongings)
    sufficient_stat_three <- colSums(belongings * matrix(data^2, nrow = length(data), ncol = ncol(belongings), byrow = FALSE))
    
    if(sum(sufficient_stat_one) != length(data)){
        stop("The sum of the first sufficient statistic does not equal node sample size")
    }
    
    if(any(sufficient_stat_three < 0)){
        stop("Third sufficient statistic is negative")
    }
    return(list("suff_stat_one" = sufficient_stat_one, 
                "suff_stat_two" = sufficient_stat_two,
                "suff_stat_three" = sufficient_stat_three)
    )
}

### test
for (i in 1:m){
    local_sufficient_stat <- calculate_node_local_sufficient_statistics(data = data_nodes[[i]], params = params_init)
    print(local_sufficient_stat)
}
```

## Functions to get updated parameter estimates

Functions:

1. calculate_global_statistics (update function)
2. estimate_parameters 

### estimate_parameters
```{r estimate_parameters}
estimate_parameters <- function(global_sufficient_statistics) {
    
    # Initialize vectors for means, standard deviations, and probabilities
    means <- numeric(length(global_sufficient_statistics$global_suff_stat_one))
    stds <- numeric(length(global_sufficient_statistics$global_suff_stat_one))
    probs <- numeric(length(global_sufficient_statistics$global_suff_stat_one))
    
    for (k in 1:length(global_sufficient_statistics$global_suff_stat_one)) {
        mean <- global_sufficient_statistics$global_suff_stat_two[k] / global_sufficient_statistics$global_suff_stat_one[k]
        var <- (global_sufficient_statistics$global_suff_stat_three[k] / global_sufficient_statistics$global_suff_stat_one[k]) - (mean^2)
        std <- sqrt(var)
        prob <- global_sufficient_statistics$global_suff_stat_one[k] / sum(global_sufficient_statistics$global_suff_stat_one)
        
        # Assign calculated values to the respective vectors
        means[k] <- mean
        stds[k] <- std
        probs[k] <- prob
    }
    
    return(list("mixing_probs" = probs,
                "means" = means,
                "stds" = stds))
}
```

### calculate_global_statistics

* gets new local sufficient statistic then updates the global sufficient statistics:
\begin{equation}
global\_stat^{new} = global\_stat^{old} + local\_stat^{new} - local\_stat^{old}
\end{equation}

```{r global_sufficient_stat}
update_global_statistics <- function(global_statistics_old,
                                     local_statistics_old,
                                     local_statistics_new,
                                     data_node,
                                     params){
    
    global_suff_stat_one <- global_statistics_old[[1]] + local_statistics_new[[1]] - local_statistics_old[[1]]
    global_suff_stat_two <- global_statistics_old[[2]] + local_statistics_new[[2]] - local_statistics_old[[2]]
    global_suff_stat_three <- global_statistics_old[[3]] + local_statistics_new[[3]] - local_statistics_old[[3]]
    
    if(sum(global_suff_stat_one) <= 0){
        stop("The sum of the first global sufficient statistics must be positive")
    }
    
    if(any(global_suff_stat_three < 0)){
        stop("Third global sufficient statistics is negative")
    }
    return(list("global_suff_stat_one" = global_suff_stat_one,
                "global_suff_stat_two" = global_suff_stat_two,
                "global_suff_stat_three" = global_suff_stat_three))
}
global_statistic_old <- list('global_suff_stat_one' = 0, 
                             'global_suff_stat_two' = 0,
                             'global_suff_stat_three' = 0)

## Test
#Have local statistics old
local_statistics_old <- lapply(1:m, function(x) calculate_node_local_sufficient_statistics(data_nodes[[x]],
                                                                                           params = params_init))
local_statistics_old
#Get global statistics old
for(i in 1:m){
    local_sufficient_stat <- calculate_node_local_sufficient_statistics(data = data_nodes[[i]], params = params_init)
    global_statistic_old$global_suff_stat_one <- global_statistic_old$global_suff_stat_one + local_sufficient_stat$suff_stat_one
    global_statistic_old$global_suff_stat_two <- global_statistic_old$global_suff_stat_two + local_sufficient_stat$suff_stat_two
    global_statistic_old$global_suff_stat_three <- global_statistic_old$global_suff_stat_three + local_sufficient_stat$suff_stat_three
}
#Get params from global statistics old
params_update <- estimate_parameters(global_statistic_old)
#Get local statistics for node 1 (new)
local_statistics_new <- calculate_node_local_sufficient_statistics(data = data_nodes[[1]], params = params_update)
local_statistics_new
#Update global statistics
global_statistic_new <- update_global_statistics(global_statistic_old,
                                                 local_statistics_old[[1]],
                                                 local_statistics_new,
                                                 data_nodes[[1]],
                                                 params_update)
global_statistic_old
local_statistics_old[[1]]
global_statistic_new
```
## Functions to check convergence
1. calculate_log_likelihood -> need to figure out the joint log-likelihood equation with global sufficient statistics / local sufficient statistics. 
2. iem_algorithm (main function) => compare difference in parameters. 

### Calculate log-likelihood (needs to be reworked)
```{r calculate_log_likelihood}
calculate_likelihood <- function(data, params) {
    mu <- params$means
    sigma <- params$stds
    weight <- params$mixing_probs
    k <- length(mu)
    n <- length(data)
    likelihood <- numeric(n)
    for (i in 1:n) {
        point_likelihood <- 0
        for (j in 1:k) {
            component_density <- weight[j] * dnorm(data[i], mean = mu[j], sd = sigma[j])
            point_likelihood <- point_likelihood + component_density
        }
        likelihood[i] <- point_likelihood
    }
    return(sum(log(likelihood)))
}

### test
likelihood_old <- calculate_likelihood(data_nodes[[1]], 
                                       params = params_init)
likelihood_update <- calculate_likelihood(data_nodes[[1]],
                                          params = params_update)
print(likelihood_old)
print(likelihood_update)
```

### IEM algorithm 
```{r iem_distributed_model}
iem_distributed_model <- function(data,
                                  initial_local_statistics,
                                  initial_global_statistics,
                                  initial_params,
                                  max_iter = 10){
    n_nodes <- length(data)
    params_old <- initial_params
    global_statistics_old <- initial_global_statistics
    local_statistics_set <- initial_local_statistics

    likelihood <- calculate_likelihood(data = unlist(data),
                                       params = params_old)
    iteration_outer <- 0
    iteration_inner <- 0
    results <- cbind(iteration_outer, iteration_inner, likelihood)
    params_diff <- 10000
    while (iteration_outer < max_iter){
        iteration_outer <- iteration_outer + 1
        iteration_inner <- 0
        for (i in 1:n_nodes){
            iteration_inner <- iteration_inner + 1
            local_statistics_old <- local_statistics_set[[i]]
            local_statistics_new <- calculate_node_local_sufficient_statistics(data = data[[i]],
                                                                                    params = params_old)
            global_statistics_new <- update_global_statistics(global_statistics_old = global_statistics_old,
                                                             local_statistics_old = local_statistics_old,
                                                             local_statistics_new = local_statistics_new,
                                                             data_node = data[[i]],
                                                             params = params_old)
            params_new <- estimate_parameters(global_sufficient_statistics = global_statistics_new)
            likelihood <- calculate_likelihood(data = unlist(data),
                                               params = params_new)
            results <- rbind(results, cbind(iteration_outer,
                                            iteration_inner,
                                            likelihood))
            params_diff <- max(abs(unlist(params_new) - unlist(params_old)))
            print(params_new)
            if (params_diff < 0.05){
                break
            }
            local_statistics_set[[i]] <- local_statistics_new
            params_old <- params_new
        }
    }
    return(list('results' = results, 'params' = params_new))
}
### Test
initial_local_statistics <- lapply(1:m, function(x) calculate_node_local_sufficient_statistics(data_nodes[[x]],
                                                                                           params = params_init))
initial_global_statistics <- list('global_suff_stat_one' = 0, 
                             'global_suff_stat_two' = 0,
                             'global_suff_stat_three' = 0)

for(i in 1:m){
    local_sufficient_stat <- calculate_node_local_sufficient_statistics(data = data_nodes[[i]], params = params_init)
    initial_global_statistics$global_suff_stat_one <- global_statistic_old$global_suff_stat_one + local_sufficient_stat$suff_stat_one
    initial_global_statistics$global_suff_stat_two <- global_statistic_old$global_suff_stat_two + local_sufficient_stat$suff_stat_two
    initial_global_statistics$global_suff_stat_three <- global_statistic_old$global_suff_stat_three + local_sufficient_stat$suff_stat_three
}
initial_params <- estimate_parameters(initial_global_statistics)
iem_model_results <- iem_distributed_model(data = data_nodes,
                                           initial_local_statistics = initial_local_statistics,
                                           initial_global_statistics = initial_global_statistics,
                                           initial_params = initial_params)

```
## Results 
### The performance of the true params
```{r true_performance}
### With true
calculate_likelihood(data=unlist(data_nodes), params = params)
plot_data_GMM(data = unlist(data_nodes), K, params)
```
### The performance of the initial params
```{r initial_performance}

### Initial
calculate_likelihood(data=unlist(data_nodes), params = params_init)
plot_data_GMM(data = unlist(data_nodes), K, params = params_init)
```
### The performance of the estimated params
```{r estimated_performance}

### Estimated
iem_model_results$results
plot_data_GMM(data = unlist(data_nodes), K, iem_model_results$params)
```