---
title: "Multivariate GMM distributed EM"
author: "Bronwyn McCall"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Getting data 

```{r libraries}
library(MASS)
library(mvtnorm)
```
```{r generating data}
#first cluster
mu1=c(0,0,0,0)
sigma1=matrix(c(3,0,0,0,0,3,0,0,0,0,3,0,0,0,0,3),ncol=4,nrow=4, byrow=TRUE)
norm1=mvrnorm(30, mu1, sigma1)
sigma1


#second cluster
mu2=c(7,7,7,7)
sigma2=sigma1
norm2=mvrnorm(30, mu2, sigma2)

#third cluster
mu3=c(3,3,3,3)
sigma3=sigma1
norm3=mvrnorm(30, mu3, sigma3)

#data
norms=rbind(norm1,norm2,norm3) #combine the 3 mixtures together
N=90 #total number of data points 
Dim=4 #number of dimensions
y=array(as.vector(norms), dim=c(N,Dim))
mixture_data=list(N=N, D=4, K=3, y=y)

means <- list(mu1, mu2, mu3)
covs <- list(sigma1, sigma2, sigma3)
mixing_probs <- list(0.33, 0.33, 1-0.33-0.33)
params <- list("means" = means, "covs" = covs, "mixing_probs" = mixing_probs)
norms
params
```

## Functions to get local sufficient statistics

Function:

1. calculate_data_belonings
2. calculate_node_local_sufficient_statistics

### Calculate_data_belongings
* Used on each worker node 
* Calculates the data points conditional component belongings based on data and current parameter estimate

```{r data_belongings}
calculate_data_belongings <- function(data, params){
    means <- params$means
    covs <- params$covs
    mixing_probs <- params$mixing_probs
    
    if (length(means) != length(covs) || length(means) != length(mixing_probs)){
        stop("Number of means, standard deviations, and weights must be equal.")
    }
    
    n_components <- length(means)
    n <- nrow(data)
    
    gamma_num <- matrix(nrow = n, ncol = n_components)
    for(i in 1:n_components){
        gamma_num[,i] <- mixing_probs[[i]] * 
            dmvnorm(data, mean = means[[i]], sigma = covs[[i]])
    }
    gamma_den <- rowSums(gamma_num)
    gamma <- gamma_num / gamma_den
    
    ### Check if data belongings sum to 1 (have grace for rounding errors).
    if (any(abs(rowSums(gamma) - 1) > .Machine$double.eps * n)){
        stop("Calculated belongings for each data point do not sum to 1.")
    }
  
    ### Check if any data belonging is negative
    if (any(gamma < 0)){
        stop("Calculated belongings contain negative values.")
    }
    return(gamma)
}

belongings <- calculate_data_belongings(data = norms, params = params)
belongings
```
### calculate_node_local_statistics
* Used on worker nodes
* For each worker node it calculates the nodes respective local sufficient statistics
```{r local_sufficient_statistics}
calculate_node_local_sufficient_statistics <- function(data, params){
    n_components <- length(params$means)
    sufficient_stat_one <- vector('list', length = n_components)
    sufficient_stat_two <- vector('list', length = n_components)
    sufficient_stat_three <- vector('list', length = n_components)
    ### Calculating the data belongings
    belongings <- calculate_data_belongings(params = params, data = data)
    
    ### Calculating the local sufficient statistics for the node
    for(i in 1:n_components){
        #### Number 1: sum of data belongings
        sufficient_stat_one[[i]] <- sum(belongings[,i])
    
        #### Number 2: sum of product (data and data belongings)
        sufficient_stat_two[[i]] <- colSums(belongings[,i] * data)
    
        #### Number 3: sum of product (data squared and data belongings)
        sufficient_stat_three[[i]] <- t(data)%*%diag(belongings[,i])%*%data
    }
    
    
    if(sum(unlist(sufficient_stat_one)) != nrow(data)){
        stop("The sum of the first sufficient statistic does not equal node sample size")
    }
    
    # if(any(unlist(sufficient_stat_three) < 0)){
    #     stop("Third sufficient statistic is negative")
    # }
    return(list("suff_stat_one" = sufficient_stat_one, 
                "suff_stat_two" = sufficient_stat_two,
                "suff_stat_three" = sufficient_stat_three)
    )
}
local_sufficient_statistics <- calculate_node_local_sufficient_statistics(data = norms, params = params)
local_sufficient_statistics
```
## Functions to get updated parameter estimates

Functions:

1. calculate_global_statistics (sum of local statistics)
2. estimate_parameters

### calculate_global_statistics

* done on manager node 
* gets the local statistics calculated on each node and sum them together to get the global statistics

```{r calculate_global_statistics}
print('Still need to implement')
```

### estimate_parameters

* done on manager node
* updates the GMM parameter estimates using the global sufficient statistics

```{r estimate_parameters}
estimate_parameters <- function(global_sufficient_statistics){
    n_components <- length(global_sufficient_statistics)
    
    means <- vector('list', length = n_components)
    covs <- vector('list', length = n_components)
    mixing_probs <- vector('list', length = n_components)
    
    for (i in 1:n_components){
        means[[i]] <- global_sufficient_statistics$suff_stat_two[[i]]/global_sufficient_statistics$suff_stat_one[[i]]
        covs[[i]] <- global_sufficient_statistics$suff_stat_three[[i]]/global_sufficient_statistics$suff_stat_one[[i]] - means[[i]]%*%t(means[[i]])
        mixing_probs <- global_sufficient_statistics$suff_stat_one[[i]]/sum(unlist(global_sufficient_statistics$suff_stat_one))
    }
    return(list("means" = means, "covs" = covs, "mixing_probs" = mixing_probs))
}

parameters <- estimate_parameters(local_sufficient_statistics)
parameters
```